---
title: "wk3 COVID data project"
author: "Serg"
date: "2023-06-08"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)


```

# COVID Info intro

This file follows the examples shown during lectures for "Data Science as a field" course.

Command output is included in most cases for educational purposes, so the whole report looks a bit untidy.

Data files are loaded from this base URL "<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/>"

individual data files:

-   "time_series_covid19_confirmed_US.csv",
-   "time_series_covid19_confirmed_global.csv",
-   "time_series_covid19_deaths_US.csv",
-   "time_series_covid19_deaths_global.csv"

And this lookup file "<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv>"

### Possible sources of bias

Bias can appear from personal beliefs of the data scientist performing
the analysis, also the way the source data was gathered, and how the
report was designed, its goals and requested analysis criteria from the
customer. All this can significantly influence the outcome. Usually bias
comes from deep beliefs, based on ancient survival mechanisms. They
usually influence someones decisions on unconscious level, and
additional steps needs to be taken to identify and prevent bias.

### Possible personal bias in the analysis

I assume my specific gender, race, previous knowledge of some geo 
specific rumors (safety, wealth) and similar beliefs, could have
impacted the way I approached this project.

### Personal bias mitigation steps taken

Knowing that some topics could be biased I took additional steps to
ensure that my analysis treats them fairly and universally. For example,
I ensured that all of data
were analysed equally, without adding any additional weights or
parameters, not relevant to the study performed.

```{r covid_init}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c(
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_deaths_global.csv",
  "time_series_covid19_confirmed_US.csv",
  "time_series_covid19_deaths_US.csv"
)
urls <- str_c(url_in, file_names)

uid_data_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

Load data from URLs

```{r import_data}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2], show_col_types = FALSE)
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

uid_data <- read_csv(uid_data_url) %>%
  select(-c(Lat,Long_,Combined_Key,iso2, iso3, Admin2, code3))
```

Clean up data

```{r tidy_global_data}
global_cases <- global_cases %>%
  pivot_longer(
    cols = -c(
      `Province/State`,
      `Country/Region`,   
      Lat,   Long
    ),
    names_to = "date",
    values_to = "cases"
  ) %>%
  select(-c(Lat ,Long))

global_deaths <- global_deaths %>%
  pivot_longer(
    cols = -c(
      `Province/State`,
      `Country/Region`,   
      Lat,   Long
    ),
    names_to = "date",
    values_to = "deaths"
  ) %>%
  select(-c(Lat ,Long))

US_cases <- us_cases %>%
  pivot_longer(
    cols = -(UID:Combined_Key),
    names_to = "date",
    values_to = "cases"
  ) %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- us_deaths %>%
  pivot_longer(
    cols = -(UID:Population),
    names_to = "date",
    values_to = "deaths"
  ) %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)
```

Transform data

```{r transform_global}
# rename cols ,for convenience & fix data format
global_data <- global_cases %>%
  full_join(global_deaths) %>%
  rename(
    Country_Region = `Country/Region`,
    Province_State = `Province/State`
  ) %>%
  mutate(date = mdy(date))
  
# remove rows with no cases
global_data <- global_data  %>%
  filter(cases > 0)

# add calculated "combined key" col
global_data <- global_data  %>%
  unite(
    "Combined_Key",
    c(Province_State, Country_Region),
    sep = ", ",
    na.rm = TRUE,
    remove = FALSE
  )

# join GLobal with population data
global_data <- global_data %>%
  left_join(uid_data, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(
    Province_State,
    Country_Region,
    date,
    cases,
    deaths,
    Population,
    Combined_Key
  )
```

```{r enrich data}
US_by_state <- US %>%
  group_by( Province_State, Country_Region, date) %>%
  summarise(cases = sum(cases), 
            deaths = sum(deaths),
            Population = sum(Population)
  ) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state %>%
  group_by( Country_Region, date) %>%
  summarise(cases = sum(cases), 
            deaths = sum(deaths),
            Population = sum(Population)
  ) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

```

```{r visualize_us}
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

```

```{r enrich_us }
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

```

Graph data of new cases / deaths for whole US

```{r visualize_us2}

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

```

Graph data of new cases / deaths for a state

```{r visualize_us3}
state <- "New York"

US_by_state %>%
  #filter(cases > 0) %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL)

US_state_totals <- US_by_state %>%
  group_by( Province_State) %>%
  summarise(cases = max(cases), 
            deaths = max(deaths),
            population = max(Population),
            cases_by_thou = 1000 * cases / population,
            deaths_by_thou = 1000 * deaths / population,
  ) %>%
  filter(cases > 0, population > 0)

print(US_state_totals %>%
  slice_min(deaths_by_thou, n = 10)
)
```

```{r modeling}
mod <- lm(deaths_by_thou ~ cases_by_thou, data = US_state_totals)

US_tot_w_pred <- US_state_totals %>%
  mutate(pred = predict(mod))

US_tot_w_pred %>%
  ggplot(  ) +
  geom_point(aes(x = cases_by_thou, y = deaths_by_thou), color = "blue") +
  geom_point(aes(x = cases_by_thou, y = pred), color = "red")
```

## session info

The above report was generated using following environment (session info):

```{r final, eval=TRUE}
sessionInfo()
```

session info finished
