---
title: "wk3-task1"
author: "Serg"
date: "2023-06-08"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The task

Use R Markdown to create your own document containing:

## a bulleted list with sub-bullets under at least one item :

-   first top level bullet
    -   sub item one
    -   sub item two
-   second top level bullet
    -   sub item one
    -   sub item two

## a numbered list

1.  first top level item
2.  second top level item

## Text formatting

some words *italicized type*

some words in **bold print**

## two levels of headers

### Header lv3

and some text

#### Header lv4

and some text again

## the results of some R code without showing the code

Simple code below

```{r first, echo=FALSE}
x <- "Hello world"
# print out the x
print(x)
```

Simple code above

## a table of text at least 3 columns wide and 3 rows long

3x3 table definition below

```{r tabwork}
df <- data.frame( 
  Name = c("Bill", "John", "Mark"),
  City = c("Boulder", "New York", "Colorado"),
  LastName = c("Smith", "Gray", "White")
)

print(df)

knitr::kable(df, format = "html")
```

and that's all for the table

## a plot (show the code that generates the plot)

Plot generation code below

```{r pressure, echo=TRUE}
plot(pressure)
```

above was the plot code

## your session info (as output from R)

my session info

```{r final, eval=TRUE}
sessionInfo()
```

session info finished
